---
title: "CCS For Procedure Codes"
---

## Intro  
CPT is a registered trademark of the American Medical Association (AMA). For basic information, please see "Decoding the Codes: CPT and RUC simplified", which can be found [here](http://www.ama-assn.org/ama/pub/physician-resources/solutions-managing-your-practice/coding-billing-insurance/cpt.page?). The codes and descriptions provide a common language to describe health care services and are used in medical billing.  

The AMA hosts a [web site](https://ocm.ama-assn.org/OCM/CPTRelativeValueSearch.do?submitbutton=accept) to enable users to look up individual CPT codes as well as reimbursement rates for both facility and non-facility sites. You can also find many web pages that will break down CPT codes for a certain specialty (for example, [psychotherapy](http://www.apapracticecentral.org/reimbursement/billing/crosswalk-codes.aspx)). But what if you have extracted thousands of procedures from your electronic health record and need some description of what the codes are?  

## Clinical Classification Software  
The [Clinical Classification Software (CCS) for services and procedures](https://www.hcup-us.ahrq.gov/toolssoftware/ccs_svcsproc/ccssvcproc.jsp) is a software tool developed as part of the [Healthcare Cost and Utilization Project](https://www.hcup-us.ahrq.gov/overview.jsp), a Federal-State-Industry partnership sponsored by the Agency for Healthcare Research and Quality. This software classifies more than 9,000 CPT codes into 244 clinically meaningful categories.  

After you download the zipped file, you'll notice the table (2015_ccs_services_procedures.csv) has CPT codes in ranges in order to adhere to copyright restrictions. However, when trying to report on many thousands of procedures, you will need a crosswalk that assigns a CCS group to each CPT code. We'll now pivot and walk through a problem that presents similar obstacles.  

## Packages  
Load data.table and [stringr](http://cran.r-project.org/web/packages/stringr/stringr.pdf):  
``` {r packages, echo=TRUE, eval=TRUE}
library(data.table)
library(stringr)
```

## Build a Custom Data Set  
Let's build a data table with three columns: range of codes, numerical class assignments, and labels for each class. We'll do this for a catalogue of products being sold in a store.  
```{r BuildData, echo=TRUE, eval=TRUE}
########## Build Code Ranges
## Set seed and generate random data
set.seed(1234)
Codes <- data.table(Start = sample(c(100:500), 10, replace = FALSE), Size = sample(c(1:5), 10, replace = TRUE), With_Letter = sample(c(1:3), 10, replace = TRUE), Letter = sample(c("A", "B", "C", "D", "E", "F", "G"), 10, replace = TRUE))

## Build end of code range
Codes[, End := Start + Size]

## Change Start and End to Characters
Codes[, Start := as.character(Start)][, End := as.character(End)]

## For items where With_Letter == 1, place letter before start and end code
Codes[With_Letter == 1, Start := paste(Letter, Start, sep = "")]
Codes[With_Letter == 1, End := paste(Letter, End, sep = "")]

## For items where With_Letter == 3, place letter at end of start and end code
Codes[With_Letter == 3, Start := paste(Start, Letter, sep = "")]
Codes[With_Letter == 3, End := paste(End, Letter, sep = "")]

## Build Ranges
Codes[, Range := paste(Start, End, sep = "-")]

## Insert single quote at beginning and end of range
Codes[, Range := paste("'", Range, "'", sep = "")]

## Remove all columns except for Ranges
Codes[, c("Start", "Size", "With_Letter", "Letter", "End") := NULL]

########## Build numberic class and class labels
Labels <- data.table(Class = c(1:10), Label = c("Soap", "Shampoo", "Conditioner", "Laundry Detergent", "Flour", "Sugar", "Deoderant", "Toothbrush", "Toothpaste", "Floss"))

########## Bind ranges and labels
data <- cbind(Codes, Labels)

########## Print data
data
```

## Overview of Problem  
From above, we have a list of ranges that each categorize items in a store. Let's say that you have a collection of receipts and on each receipt is an item number that falls within the ranges. If you want to easily classify each item, then you need a one-to-one crosswalk between item codes and categories.  

We'll break this problem into two parts. First, we'll handle cases where the ranges of codes don't have letters. Second, we'll handle cases where the range of codes have a letter at the beginning or a letter at the end.  

Finally, let's say that each code must be 5 characters long. For example, we must append leading zeros to the code for item number A62, so that it appears as "A0062".  

## Step 1: Handle Cases Without Letters  
```{r NoLetters_Fxn, echo=TRUE, eval=TRUE}
Crosswalk_NoLetters <- function(data){
    ########## Clean Data
    ## Subset data to only contain ranges without letters
    noletters <- data[!(grep("[A-Z]+", Range))]
    
    ## Remove single quote
    noletters[, Range := gsub("'", "", Range)]
    
    ## Split Range into Start and End
    noletters[, c("Start", "End") := tstrsplit(Range, "-", fixed=TRUE)]
    
    ## Convert class to numeric
    noletters[, Start := as.numeric(Start)][, End := as.numeric(End)]
    
    ########## Set up writing out
    ## Create new table to write codes to
    output <- data.table(Code=rep(0,100000), Class=rep(0,100000), Label=rep("empty",100000))
    
    ## Set line in output to write to
    out_line <- 1L
    
    ########## Read through every line in data set
    for(i in 1:nrow(noletters)){
        ## Calculate the number of times to repeat rows
        Reps <- (noletters[i,End] - noletters[i,Start] + 1)
        
        ## And loop through to write new lines for each code
        for(j in 1:Reps){
            ## The j index writes (Reps) new rows, where the value in the code column of each incremental row increases by 1
            set(output,out_line + (j - 1L),1L, noletters[i,Start] + (j - 1L))
            set(output,out_line + (j - 1L),2L, noletters[i,Class])
            set(output,out_line + (j - 1L),3L, noletters[i,Label])
        }
        ## Bump up the line to print from by the number of Replicate rows written
        out_line <- as.integer(out_line + Reps)
    }
    
    ########## Append leading zeros
    ## Keep output that contains data
    output <- output[Code != 0]
    
    ## Convert code to a strong
    output[, Code := as.character(Code)]
    
    ## Count number of leading zeros to insert
    output[, Leading_Zeros := as.integer(5 - nchar(Code))]
    
    ## Fill leading zeros in Code column for the rows that require it 
    for (i in which(output$Leading_Zeros != 0))
        set(output, i, 1L, paste(c(rep("0", times = output[i, Leading_Zeros]), output[i, Code]), collapse = ""))

    ## Drop leading zeros column
    output[, Leading_Zeros := NULL]
    
    ########## Return output that contains data
    return(output)
}
```

## Step 2: Handle Cases With Letters  
```{r WithLetters_Fxn, eval=TRUE,echo=TRUE}
Crosswalk_Letters <- function(data){
    ########## Clean data
    ## Subset data to only contain ranges without letters
    withletters <- data[grep("[A-Z]+", Range)]
    
    ## Remove single quote
    withletters[, Range := gsub("'", "", Range)]
    
    ## Pull letter at beginning of line
    withletters[grep("^[A-Z]", Range), At_Start := substr(Range, 1, 1)]

    ## Pull letter at end of line
    withletters[grep("[A-Z]$", Range), At_End := str_sub(Range, -1)]

    ## Remove letters from range
    withletters[, Range := gsub("[A-Z]", "", Range)]
    
    ## Split Range into Start and End
    withletters[, c("Start", "End") := tstrsplit(Range, "-", fixed=TRUE)]
    
    ## Convert class to numeric
    withletters[, Start := as.numeric(Start)][, End := as.numeric(End)]
    
    ########## Set up writing out
    ## Create new table to write codes to
    output <- data.table(Code=rep(0,100000), Class=rep(0,100000), Label=rep("empty",100000), At_Start = rep("empty",100000), At_End = rep("empty",100000))
    
    ## Set line in output to write to
    out_line <- 1L
    
    ########## Read through every line in data set
    for(i in 1:nrow(withletters)){
        ## Calculate the number of times to repeat rows
        Reps <- (withletters[i,End] - withletters[i,Start] + 1)
        
        ## And loop through to write new lines for each code
        for(j in 1:Reps){
            ## The j index writes (Reps) new rows, where the value in the code column of each incremental row increases by 1
            set(output,out_line + (j - 1L),1L, withletters[i,Start] + (j - 1L))
            set(output,out_line + (j - 1L),2L, withletters[i,Class])
            set(output,out_line + (j - 1L),3L, withletters[i,Label])
            set(output,out_line + (j - 1L),4L, withletters[i,At_Start])
            set(output,out_line + (j - 1L),5L, withletters[i,At_End])
        }
        ## Bump up the line to print from by the number of Replicate rows written
        out_line <- as.integer(out_line + Reps)
    }
    
    ########## Drop output without data
    output <- output[Code != 0]
    
    ########## Append leading zeros
    ## Convert code back to string
    output[, Code := as.character(Code)]
    
    ## Count number of leading zeros to insert
    output[, Leading_Zeros := as.integer(4 - nchar(Code))]
    
    ## Fill leading zeros in Code column for the rows that require it 
    for (i in which(output$Leading_Zeros != 0))
        set(output, i, 1L, paste(c(rep("0", times = output[i, Leading_Zeros]), output[i, Code]), collapse = ""))
    
    ########## Replace letter at beginning or end
    ## Add letter at end
    output[!is.na(At_End), Code := paste(Code, At_End, sep = "")]

    ## Add letter at start
    output[!is.na(At_Start), Code := paste(At_Start, Code, sep = "")]

    ########## Remove extra columns
    output[, c("At_Start", "At_End", "Leading_Zeros") := NULL]
    
    ## Return output
    return(output)
}
```

## Step 3: Run functions and append results    
```{r AppendResults, eval=TRUE,echo=TRUE}
crosswalk <- rbind(Crosswalk_NoLetters(data), Crosswalk_Letters(data))
print(crosswalk)
```